Endogenous Macrodynamics in Algorithmic Recourse

### 2022-08-09 dev/notebooks/gravitational_generator.qmd
extended methodology to include Gravitational generator

```{julia}
Λ₂ = [0.1, 1, 5]
counterfactuals = []
generators = []
for λ₂ ∈ Λ₂  
    λ = [0.1, λ₂]
    generator = GravitationalGenerator(λ=λ)
    generators = vcat(generators..., generator)
    counterfactuals = vcat(
      counterfactuals...,
      generate_counterfactual(x, target, counterfactual_data, M, generator; convergence=:strict, T=T)
    )
end
```

```{julia}
theme(:wong)
T_ = 500
plts = []
for i ∈ 1:length(Λ₂)
    λ₂ = Λ₂[i]
    counterfactual = counterfactuals[i]  
    plt = plot(counterfactual, plot_up_to=minimum([T,T_]), title="λ₂=$(λ₂)")
    x1,x2 = generators[i].centroid[1], generators[i].centroid[2]
    scatter!(plt, [x1], [x2], colour=:purple, label="attractor")
    plts = vcat(plts..., plt)
end
plt = plot(plts..., size=(1200,300), layout=(1,3))
savefig(plt, joinpath(www_path,"gravitational_generator_strict.png"))
```

@@

```{julia}

```

>>>>

```{julia}
Λ₂ = [0.1, 1, 5]
counterfactuals_strict = []
generators = []
for λ₂ ∈ Λ₂  
    λ = [0.1, λ₂]
    generator = GravitationalGenerator(λ=λ)
    generators = vcat(generators..., generator)
    counterfactuals_strict = vcat(
      counterfactuals_strict...,
      generate_counterfactual(x, target, counterfactual_data, M, generator; convergence=:strict, T=T)
    )
end
```

```{julia}
theme(:wong)
T_ = 500
plts = []
for i ∈ 1:length(Λ₂)
    λ₂ = Λ₂[i]
    counterfactual = counterfactuals_strict[i]  
    plt = plot(counterfactual, plot_up_to=minimum([T,T_]), title="λ₂=$(λ₂)")
    x1,x2 = generators[i].centroid[1], generators[i].centroid[2]
    scatter!(plt, [x1], [x2], colour=:purple, label="attractor")
    plts = vcat(plts..., plt)
end
plt = plot(plts..., size=(1200,300), layout=(1,3))
savefig(plt, joinpath(www_path,"gravitational_generator_strict.png"))
```

@@

### Comparison - simple vs. strict convergence

```{julia}
idx = 2
x1,x2 = generators[idx].centroid[1], generators[idx].centroid[2]
plt1 = plot(counterfactuals[idx], plot_up_to=minimum([T,T_]), title="Simple")
scatter!(plt1, [x1], [x2], colour=:purple, label="attractor")
plt2 = plot(counterfactuals_strict[idx], plot_up_to=minimum([T,T_]), title="Strict")
scatter!(plt2, [x1], [x2], colour=:purple, label="attractor")
plt = plot(plt1, plt2, size=(850,350), layout=(1,2))
savefig(plt, joinpath(www_path,"gravitational_generator_comparison.png"))
```

### 2022-08-09 dev/paper/paper.tex
extended methodology to include Gravitational generator

\begin{document}
\maketitle
\begin{abstract}
Existing work on Counterfactual Explanations (CE) and Algorithmic
Recourse (AR) has largely been limited to the static setting: given some
classifier we are interested in finding close, actionable, realistic,
sparse, diverse and ideally causally founded counterfactuals. The
ability of CE to handle dynamics like data and model drift remains a
largely unexplored research challenge at this point. Only one recent
work considers the implications of exogenous domain and model shifts.
This project instead focuses on endogenous dynamics, that is shifts that
occur when AR is actually implemented by a proportion of individuals.
Early findings suggest that the involved shifts may be large with
important implications on the validity of AR and the overall
characteristics of the sample population.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, sharp corners, enhanced, borderline west={3pt}{0pt}{shadecolor}, breakable, boxrule=0pt, frame hidden]}{\end{tcolorbox}}\fi

@@

In this work we have chosen to focus on a number of gradient-based
counterfactual generators to investigate the endogenous dynamics we
introduced in Section~\ref{sec-intro}. Gradient-based counterfactual
search is well-suited for differentiable black-box models like deep
neural networks. We can restate Equation~\ref{eq-solution} in a more
general form that encompasses most gradient-based approaches to
counterfactual search:

@@

@@

\hypertarget{sec-limit}{%
\subsection{Limitations and Future Work}\label{sec-limit}}

@@

\hypertarget{data}{%
\subsubsection{Data}\label{data}}

The number of real-world datasets considered in this work is limited,
constrained by the fact that at the time of writing
\texttt{CounterfactualExplanations.jl} only supported continuous
features.

@@

>>>>

\begin{document}
\maketitle
\begin{abstract}
Existing work on Counterfactual Explanations (CE) and Algorithmic
Recourse (AR) has largely been limited to the static setting: given some
classifier we are interested in finding close, actionable, realistic,
sparse, diverse and ideally causally founded counterfactuals. The
ability of CE to handle dynamics like data and model drift remains a
largely unexplored research challenge at this point. Only one recent
work considers the implications of exogenous domain and model shifts.
This project instead focuses on endogenous dynamics, that is shifts that
occur when AR is actually implemented by a proportion of individuals.
Early findings suggest that the involved shifts may be large with
important implications on the validity of AR and the overall
characteristics of the sample population.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, interior hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, boxrule=0pt, sharp corners, frame hidden]}{\end{tcolorbox}}\fi

@@

In this work we have chosen to focus on a number of gradient-based
counterfactual generators to investigate the endogenous dynamics we
introduced in Section~\ref{sec-intro}. Gradient-based counterfactual
search is well-suited for differentiable black-box models like deep
neural networks.

\hypertarget{from-individual-recourse}{%
\subsubsection{From individual recourse
\ldots{}}\label{from-individual-recourse}}

We can restate Equation~\ref{eq-solution} in a more general form that
encompasses most gradient-based approaches to counterfactual search:

@@

\hypertarget{towards-collective-recourse}{%
\subsubsection{\ldots{} towards collective
recourse}\label{towards-collective-recourse}}

To explicitly address the notion that individual recourse may affect the
outcome and prospect of other individuals, we propose to extend
Equation~\ref{eq-general} as follows:

\begin{equation}\protect\hypertarget{eq-collective}{}{
\begin{aligned}
\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}}  \sum_{k=1}^{K} {\ell(M(f(s_k^\prime)),t)} \\ &+ \lambda_1 {h(f(s_k^\prime)) } + \lambda_2 {g(f(s_k^\prime))}  
\end{aligned}
}\label{eq-collective}\end{equation}

Here \(h(f(s_k^\prime))\) denotes the proxy for private costs faced by
the individual as before and \(\lambda_1\) governs to what extent that
private cost ought to be penalised. The newly introduced term
\(g(f(s_k^\prime))\) is meant to capture and address social costs
incurred by the collective of individuals in response to changes in
\(\mathbf{s}^\prime\). The trade-off between private and social costs is
determined by the ratio between \(\lambda_1\) and \(\lambda_2\). As with
individual recourse, the exact choice of \(g(\cdot)\) is not obvious,
nor do we intend to provide a definite answer in this work, if such even
exists. A straight-forward choice simply extends the baseline approach
by \protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}}: instead
of only penalizing the distance of the individuals' counterfactual to
its factual, we propose penalizing its distance to some sensible point
in the target domain, for example the sample average:
\(\bar{\mathbf{x}}\). For such a recourse objective, higher choices of
\(\lambda_2\) relative to \(\lambda_1\) will lead counterfactuals to
gravitate towards the specified point in the target domain. In the
remainder of this paper we will therefore refer to this approach as
\textbf{Gravitational} generator, when we investigate its potential
usefulness for mitigating endongenous macrodynamics\footnote{Note that
  despite the naming convention our goal here is not to provide yet
  another counterfactual generator, but merely investigate the most
  simple penalty we can think of with respect to its effectiveness.}.

\hypertarget{a-note-on-convergence}{%
\subsubsection{A note on convergence}\label{a-note-on-convergence}}

For this simple mitigating strategy underlying the Gravitational
generator to work as expected, one needs to ensure that counterfactual
search continues, even after a predetermined threshold probability
\(\gamma\) has potentially already been reached.
Figure~\ref{fig-convergence} illustrates this distinction: if one
chooses to terminate search once the desired threshold is reached (left
panel) the gravitational pull towards \(\bar{\mathbf{x}}\) is never
actually satisfied (compare to right panel). More generally, if
convergence is defined simply in terms of flipping the predicted label
with some desired degree of confidence, this corresponds to essentially
ignoring any parts of the counterfactual search objective that do not
involve \(\ell(M(f(s_k^\prime)),t)\) beyond that point. While this may
be appropriate for some applications, in general this seems like an odd
convention. Since we nonetheless seen convergence specified simply in
terms of reaching the threshold probability in some places\footnote{\protect\hyperlink{ref-joshi2019towards}{{[}8{]}}
  define convergence of Algorithm 1 in this way. The implementation of
  \protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}} in CARLA is
  also defined in this way.}, we thought it worth making this
distinction explicit.

{\centering \includegraphics[width=0.45\textwidth,height=\textheight]{www/gravitational_generator_comparison.png}

\caption{\label{fig-convergence}Comparison of counterfactual search
outcome with simple (left) and strict convergence (right).}

@@

\hypertarget{sec-limit}{%
\section{Limitations and Future Work}\label{sec-limit}}

While we believe that this work constitutes a valuable starting point
for addressing existing issues in algorithmic recourse from a fresh
perspective, we are aware of several of its limitations. In the
following we highlight some of these limitations and point to avenues
for future research.

\hypertarget{experimental-setup}{%
\subsection{Experimental Setup}\label{experimental-setup}}

@@

\hypertarget{data}{%
\subsection{Data}\label{data}}

Largely in line with the existing literature on algorithmic recourse, we
have limited our analysis of real-world data to three commonly used
benchmark datasets that involve binary prediction tasks. Future work may
benefit from including novel datasets or extending the analysis to
multi-class or regression problems, the latter arguably representing the
most common objective in Finance and Economics. It is also worth
mentioning that the use of real-world datasets considered in this work
is constrained by the fact that at the time of writing
\texttt{CounterfactualExplanations.jl} only supports continuous
features, at least of some of the counterfactual generators considered
here. The fact that we therefore had to discard discrete features led to
relatively poor initial performance of our classifiers in some cases.
While this is indeed a limitation we intend to address in future and
derivative work, our findings with respect to endogenous macrodynamics
do not hinge on strong classifier performance.

\hypertarget{classifiers}{%
\subsection{Classifiers}\label{classifiers}}

For reasons stated earlier we have limited our analysis to
differentiable linear and non-linear classifiers, in particular logistic
regression and deep neural networks. While these sorts of classifiers
have also typically been analyzed in the existing literature on
counterfactual explanations and algorithmic recourse, they represent
only a subset of popular machine learning models employed in practice -
both black-box and glass-box. Despite the success and popularity of deep
learning in the context of high-dimensional data such as image, audio
and video, empirical evidence suggests that other models such as boosted
decision trees may have an edge when it comes to lower-dimensional
tabular datasets, such as the ones considered here
\protect\hyperlink{ref-grinsztajn2022tree}{{[}28{]}}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-borisov2021deep}{}}%
\CSLLeftMargin{{[}27{]} }%
\CSLRightInline{V. Borisov, T. Leemann, K. Seßler, J. Haug, M.
Pawelczyk, and G. Kasneci, {``Deep neural networks and tabular data: A
survey,''} \emph{arXiv preprint arXiv:2110.01889}, 2021.}

\leavevmode\vadjust pre{\hypertarget{ref-grinsztajn2022tree}{}}%
\CSLLeftMargin{{[}28{]} }%
\CSLRightInline{L. Grinsztajn, E. Oyallon, and G. Varoquaux, {``Why do
tree-based models still outperform deep learning on tabular data?''}
\emph{arXiv preprint arXiv:2207.08815}, 2022.}

### 2022-08-09 dev/paper/sections/methodology.qmd
extended methodology to include Gravitational generator

In this work we have chosen to focus on a number of gradient-based counterfactual generators to investigate the endogenous dynamics we introduced in @sec-intro. Gradient-based counterfactual search is well-suited for differentiable black-box models like deep neural networks. We can restate @eq-solution in a more general form that encompasses most gradient-based approaches to counterfactual search:

@@

>>>>

In this work we have chosen to focus on a number of gradient-based counterfactual generators to investigate the endogenous dynamics we introduced in @sec-intro. Gradient-based counterfactual search is well-suited for differentiable black-box models like deep neural networks.

### From individual recourse ...

We can restate @eq-solution in a more general form that encompasses most gradient-based approaches to counterfactual search:

@@

### ... towards collective recourse

To explicitly address the notion that individual recourse may affect the outcome and prospect of other individuals, we propose to extend @eq-general as follows:

$$
\begin{aligned}
\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}}  \sum_{k=1}^{K} {\ell(M(f(s_k^\prime)),t)} \\ &+ \lambda_1 {h(f(s_k^\prime)) } + \lambda_2 {g(f(s_k^\prime))}  
\end{aligned}
$$ {#eq-collective}

Here $h(f(s_k^\prime))$ denotes the proxy for private costs faced by the individual as before and $\lambda_1$ governs to what extent that private cost ought to be penalised. The newly introduced term $g(f(s_k^\prime))$ is meant to capture and address social costs incurred by the collective of individuals in response to changes in $\mathbf{s}^\prime$. The trade-off between private and social costs is determined by the ratio between $\lambda_1$ and $\lambda_2$. As with individual recourse, the exact choice of $g(\cdot)$ is not obvious, nor do we intend to provide a definite answer in this work, if such even exists. A straight-forward choice simply extends the baseline approach by @wachter2017counterfactual: instead of only penalizing the distance of the individuals' counterfactual to its factual, we propose penalizing its distance to some sensible point in the target domain, for example the sample average: $\bar{\mathbf{x}}$. For such a recourse objective, higher choices of $\lambda_2$ relative to $\lambda_1$ will lead counterfactuals to gravitate towards the specified point in the target domain. In the remainder of this paper we will therefore refer to this approach as **Gravitational** generator, when we investigate its potential usefulness for mitigating endongenous macrodynamics^[Note that despite the naming convention our goal here is not to provide yet another counterfactual generator, but merely investigate the most simple penalty we can think of with respect to its effectiveness.].

### A note on convergence

For this simple mitigating strategy underlying the Gravitational generator to work as expected, one needs to ensure that counterfactual search continues, even after a predetermined threshold probability $\gamma$ has potentially already been reached. @fig-convergence illustrates this distinction: if one chooses to terminate search once the desired threshold is reached (left panel) the gravitational pull towards $\bar{\mathbf{x}}$ is never actually satisfied (compare to right panel). More generally, if convergence is defined simply in terms of flipping the predicted label with some desired degree of confidence, this corresponds to essentially ignoring any parts of the counterfactual search objective that do not involve $\ell(M(f(s_k^\prime)),t)$ beyond that point. While this may be appropriate for some applications, in general this seems like an odd convention. Since we nonetheless seen convergence specified simply in terms of reaching the threshold probability in some places^[@joshi2019towards define convergence of Algorithm 1 in this way. The implementation of @wachter2017counterfactual in CARLA is also defined in this way.], we thought it worth making this distinction explicit.

![Comparison of counterfactual search outcome with simple (left) and strict convergence (right).](www/gravitational_generator_comparison.png){#fig-convergence fig.pos="h" width=45%}
